---
title: "Clean and Explore VZinfo Data"
author: "Laurens Bogaardt"
date: "2023-09-21"
output:
  html_document:
    df_print: paged
    fig_width: 10
    fig.align: center
    toc: true
    toc_depth: 4
    toc_float: true
    theme: united
    code_folding: show
bibliography: ../A-Model-of-Individual-BMI-Trajectories.bib
link-citations: yes
---

<style>
body{text-align: justify}
</style>

# Introduction

The purpose of this document is to clean and explore the historical BMI data of VZinfo. Historical data about BMI were assessed by Statistics Netherlands and the National Institute for Public Health and the Environment [@CBS2021]. We use the prevalences of overweight and obesity for adults between 1990 and 2021 which were standardised to the demography of the Netherlands in 2021 [@VZinfo2021]. Due to this standardisation step, any trend we extract from these data will mostly be indicative of changes in lifestyle and its influence on BMI rather than also including trends in demography.

Before we begin, we need to load the package _tidyverse_ [@tidyverse2019].

```{r results = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)
```

# Load Input

The historical BMI data of VZinfo comes in two separate datasets containing the prevalences of overweight and of obesity over time. Let's first load the dataset containing the prevalences of overweight over time which contains one column indicating the year, another for the prevalences of males and one more for the prevalences of females.

```{r}
trend.volwassenen.met.overgewicht <- read_delim(
  file = "../Input-Data/VZinfo/trend-volwassenen-met-ov.csv",
  delim = ";",
  col_types = cols_only(
    "Category" = col_integer(),
    "mannen gestand.*" = col_double(),
    "vrouwen gestand.*" = col_double()
  )
)
trend.volwassenen.met.overgewicht
```

Let's now load the dataset containing the prevalences of obesity over time.

```{r}
trend.volwassenen.met.obesitas <- read_delim(
  file = "../Input-Data/VZinfo/trend-volwassenen-met-ob.csv",
  delim = ";",
  col_types = cols_only(
    "Category" = col_integer(),
    "mannen gestand.*" = col_double(),
    "vrouwen gestand.*" = col_double()
  )
)
trend.volwassenen.met.obesitas
```

# Clean Data

We clean the data by first pivoting the datasets such that we obtain a one column with the prevalences and a new column indicating whether this applies to males of females. Both datasets are combined and again pivoted to create a new column indicating whether the prevalences apply to overweight or obesity.

```{r}
bmi.by.year.and.sex <- trend.volwassenen.met.overgewicht %>%
  pivot_longer(
    cols = !Category,
    names_pattern = "(.*) .*",
    names_to = "sex",
    values_to = "prevalence_[25,Inf)"
  ) %>%
  full_join(
    trend.volwassenen.met.obesitas %>%
      pivot_longer(
        cols = !Category,
        names_pattern = "(.*) .*",
        names_to = "sex",
        values_to = "prevalence_[30,Inf)"
      ),
    by = c("Category", "sex")
  ) %>%
  mutate(
    sex = recode_factor(
      sex,
      "mannen" = 0,
      "vrouwen" = 1
    ),
    year = Category,
    `prevalence_[25,30)` = round((`prevalence_[25,Inf)` - `prevalence_[30,Inf)`) / 100, 3),
    `prevalence_[30,Inf)` = round(`prevalence_[30,Inf)` / 100, 3)
  ) %>%
  pivot_longer(
    cols = c("prevalence_[25,30)", "prevalence_[30,Inf)"),
    names_pattern = "prevalence_(.*)",
    names_to = "bmi",
    values_to = "prevalence"
  ) %>%
  select(year, sex, bmi, prevalence) %>%
  arrange(year, sex, bmi, prevalence)
bmi.by.year.and.sex
```

# Write Output

Finally, we write the cleaned data of VZinfo to a CSV file.

```{r}
write_csv(
  x = bmi.by.year.and.sex,
  file = "../Output-Data/VZinfo.csv"
)
```

# References

<div id="refs"></div>
