---
title: "Clean and Explore Gezondheidsmonitor Volwassenen"
author: "Laurens Bogaardt"
date: "2023-02-01"
output:
  html_document:
    df_print: paged
    fig_width: 10
    fig.align: center
    toc: true
    toc_depth: 4
    toc_float: true
    theme: united
    code_folding: show
bibliography: R:\\Projecten\\V02001214 KV2.1.B CZmodellering\\6. Producten\\2022-08-15 - Mathematical Medicine and Biology - A Model of Individual BMI Trajectories\\Submitted\\Bibliography.bib
link-citations: yes
---

<style>
body{text-align: justify}
</style>

# Introduction

The purpose of this document is to clean and explore the Gezondheidsmonitor dataset. This is a Dutch cross-sectional dataset based on a large, health-related questionnaire administered by the Community Health Services, Statistics Netherlands and the National Institute for Public Health and the Environment [@GGD2012]. The dataset contains self-reported weight and height measurements, rounded to nearest kilogram and centimetre. The included survey weights allow for representative analyses of the Netherlands.

Before we begin, we need to load the packages _tidyverse_, _haven_ and _vtable_ [@tidyverse2019; @haven2021; @vtable2021].

```{r results = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)
library(haven)
library(vtable)
```

# Load Input

[Placeholder for Text]

```{r}
population.by.year.sex.education.and.age <- read_csv(
  file = "../CBS/Demography/Population Size by Date, Sex, Education and Age.csv",
  col_type = cols_only(
    date = col_date(),
    sex = col_factor(),
    education = col_factor(),
    age_lower = col_double(),
    age_upper = col_double(),
    population.size = col_integer()
  )
)
population.by.year.sex.education.and.age
```

Let's load the Gezondheidsmonitor dataset. The survey was first conducted in September, October and November of 2012, and was repeated in 2016 and in 2020. We will combine all three datasets here. Throughout the document, we will limit the output in each step. This is because the Gezondheidsmonitor data are not open source. To request access to the data, please visit [www.monitorgezondheid.nl](https://www.monitorgezondheid.nl).

```{r}
gezondheidsmonitor.data <- read_sav(
  file = "R:/Projecten/V010041 Monitor/DataUitgifte/Data Gezondheidsmonitor Volwassenen en Ouderen 2012/2020_03_31_GM2012_Totaalbestand GGD en CBS (N=387195).sav",
  col_select = c("Surveynummer", "weegfactor3", "geslacht", "LFT0109", "opleiding_def", "AGGWS201", "LFALA201", "LFALA202", "LFALS218", "LFRKA205", "LFRKA206", "LFRKA207")
) %>%
  zap_labels() %>%
  rename(
    sex = geslacht,
    education = opleiding_def,
    weight = weegfactor3,
    age = LFT0109, # Is geboortejaar beschikbaar
    bmi = AGGWS201,
    drinker_last_year = LFALA201,
    exdrinker = LFALA202,
    #neverdrinker = LFALB207, #does not exist in the supplied dataset somehow
    glasses_per_week = LFALS218,
    roker = LFRKA205,
    exroker = LFRKA206,
    nooitroker = LFRKA207
  ) %>%
  mutate(year = 2012) %>%
  bind_rows(
    read_sav(
      file = "R:/Projecten/V010041 Monitor/DataUitgifte/Data Gezondheidsmonitor Volwassenen en Ouderen 2016/2018_06_20_Totaalbestand GGD en CBS (N=457153) Landelijk.sav",
      col_select = c("Surveynummer", "ewGGD", "geslacht", "LFT0109", "Opleiding_samind", "AGGWS201", "LFALA201", "LFALA202", "alc_week","LFRKA205", "LFRKA206", "LFRKA207")
    ) %>%
      zap_labels() %>%
      rename(
        sex = geslacht,
        education = Opleiding_samind,
        weight = ewGGD,
        age = LFT0109,
        bmi = AGGWS201,
        drinker_last_year = LFALA201,
        exdrinker = LFALA202,
        #neverdrinker = LFALB207, #does not exist in the supplied dataset somehow
        glasses_per_week = alc_week,
        roker = LFRKA205,
        exroker = LFRKA206,
        nooitroker = LFRKA207
      ) %>%
      mutate(year = 2016)
  ) %>%
  bind_rows(
    read_sav(
      file = "R:/Projecten/V010041 Monitor/DataUitgifte/Data Gezondheidsmonitor Volwassenen en Ouderen 2020/2022_2_28Totaalbestand GGD en CBS (N=539.902) Landelijk.sav",
      col_select = c("Surveynummer", "ewGGDkern", "Geslacht", "LFT0109", "Opleiding_samind", "AGGWS201", "LFALA217", "LFALA218", "alc_week", "LFRKA205", "LFRKA206", "LFRKA207")
    ) %>%
      zap_labels() %>%
      rename(
        sex = Geslacht,
        education = Opleiding_samind,
        weight = ewGGDkern,
        age = LFT0109,
        bmi = AGGWS201,
        drinker_last_year = LFALA217,
        exdrinker = LFALA218,
        #neverdrinker = LFALB207, #does not exist in the supplied dataset somehow
        glasses_per_week = alc_week,
        roker = LFRKA205,
        exroker = LFRKA206,
        nooitroker = LFRKA207
      ) %>%
      mutate(year = 2020)
  )
```

# Clean Data

We will examine each variable in the dataset and clean the data when necessary.

## Survey Number

The Gezondheidsmonitor dataset is a combination of a questionnaire administered by the Community Health Services and one administered by Statistics Netherlands. The variable _Surveynummer_ indicates which questionnaire the participant participated in.

```{r}
gezondheidsmonitor.data %>%
  count(Surveynummer)
```

We are only interested in the data gathered by the Community Health Services, so we will filter our dataset. The variable _Surveynummer_ is subsequently no longer needed.

```{r}
gezondheidsmonitor.data <- gezondheidsmonitor.data %>%
  filter(Surveynummer == 1) %>%
  select(-Surveynummer)
```

## Weight

The survey weights included in the dataset allow for representative analyses of the Netherlands. Let's check if any of these are missing and whether they are all positive numbers.

```{r}
gezondheidsmonitor.data %>%
  summarise(
    min = min(weight),
    max= max(weight),
    na = sum(is.na(weight)),
    zero = sum(weight == 0)
  )
```

Let's remove the participants who have zero weight.

```{r}
gezondheidsmonitor.data <- gezondheidsmonitor.data %>%
  filter(weight > 0)
```

## Sex

We verify that there are only two sexes and immediately check whether there are any missing values.

```{r}
gezondheidsmonitor.data %>%
  count(sex)
```

Let's recode the variable. A _0_ will encode males, whereas females are encoded by a _1_.

```{r}
gezondheidsmonitor.data <- gezondheidsmonitor.data %>%
  mutate(sex = factor(sex - 1))
```

## Education

Let's now check whether any participant is missing their level of education.

```{r}
gezondheidsmonitor.data %>%
  count(education)
```

We will remove these participants from the dataset.

```{r}
gezondheidsmonitor.data <- gezondheidsmonitor.data %>%
  filter(!is.na(education))
```

Let's now recode the variable. Education was measured as the highest level reached and categorised into three levels. The lowest level, encoded by _1_, applies to people with intermediate secondary education or less, the medium level, encoded by _2_, aggregates higher secondary- and intermediate vocational education and highest level, encoded by _3_, includes higher vocational education and university.

```{r}
gezondheidsmonitor.data <- gezondheidsmonitor.data %>%
  mutate(
    education = ifelse(education == 1, education, education - 1), #recode the education levels. 
    # original:  laag (LO), midden1 (MAVO, LBO), midden2 (HAVO, VWO, MBO) en hoog (HBO, WO)
    # source: https://monitorgezondheid.nl/sites/default/files/2021-06/20160914%20Opbouw%20en%20instructie%20totaalbestand%20Gezondheidsmonitor%202012.pdf
    education = factor(education, levels = c(1, 2, 3), ordered = TRUE)
  )
```

## Age

Let's check that there are only adults in the dataset. We also verify that there are no missing values.

```{r}
gezondheidsmonitor.data %>%
  summarise(
    min = min(age),
    max = max(age),
    na = sum(is.na(age))
  )
```

# Verify Representativeness

[Placeholder for Text]

```{r}
gezondheidsmonitor.data %>%
  filter(
    age >= 25,
    age < 75
  ) %>%
  group_by(year, sex) %>%
  summarise(
    population.size = sum(weight),
    .groups = "drop_last"
  ) %>%
  mutate(
    prevalence = population.size / sum(population.size)
  ) %>%
  ungroup()
```

[Placeholder for Text]

```{r}
population.by.year.sex.education.and.age %>%
  filter(
    year(date) %in% c(2014, 2016, 2020),
    age_lower >= 25,
    age_upper <= 75
  ) %>%
  group_by(year = year(date), sex) %>%
  summarise(
    population.size = sum(population.size),
    .groups = "drop_last"
  ) %>%
  mutate(
    prevalence = population.size / sum(population.size)
  ) %>%
  ungroup()
```

[Placeholder for Text]

```{r}
gezondheidsmonitor.data %>%
  filter(
    age >= 25,
    age < 75
  ) %>%
  group_by(year, education) %>%
  summarise(
    population.size = sum(weight),
    .groups = "drop_last"
  ) %>%
  mutate(
    prevalence = population.size / sum(population.size)
  ) %>%
  ungroup()
```

[Placeholder for Text]

```{r}
population.by.year.sex.education.and.age %>%
  filter(
    year(date) %in% c(2014, 2016, 2020),
    age_lower >= 25,
    age_upper <= 75
  ) %>%
  group_by(year = year(date), education) %>%
  summarise(
    population.size = sum(population.size),
    .groups = "drop_last"
  ) %>%
  mutate(
    prevalence = population.size / sum(population.size)
  ) %>%
  ungroup() # Niet in weegmodel
```

[Placeholder for Text]

```{r}
gezondheidsmonitor.data %>%
  filter(
    age >= 25,
    age < 75
  ) %>%
  group_by(
    year,
    age = cut(age, c(25, 35, 45, 55, 65, 75), include.lowest = TRUE, right = FALSE)
  ) %>%
  summarise(
    population.size = sum(weight),
    .groups = "drop_last"
  ) %>%
  mutate(
    prevalence = population.size / sum(population.size)
  ) %>%
  ungroup()
```

[Placeholder for Text]

```{r}
population.by.year.sex.education.and.age %>%
  filter(
    year(date) %in% c(2014, 2016, 2020),
    age_lower >= 25,
    age_upper <= 75
  ) %>%
  group_by(
    year = year(date),
    age = cut(age_lower, c(25, 35, 45, 55, 65, 75), include.lowest = TRUE, right = FALSE)
  ) %>%
  summarise(
    population.size = sum(population.size),
    .groups = "drop_last"
  ) %>%
  mutate(
    prevalence = population.size / sum(population.size)
  ) %>%
  ungroup()
```

[Placeholder for Text]

# Clean BMI

In most of their analyses, Statistics Netherlands removes outlying BMI values whereby it uses $14 kg / m^2$ and $50 kg / m^2$ as the lower and upper limits. Let's check whether we have such outliers in our data and if there are any missing values.

```{r}
gezondheidsmonitor.data %>%
  summarise(
    min = min(bmi, na.rm = TRUE),
    max = max(bmi, na.rm = TRUE),
    na = sum(is.na(bmi)),
    below = sum(bmi < 14, na.rm = TRUE),
    above = sum(bmi > 50, na.rm = TRUE)
  )
```

Let's remove any BMI value larger than 50.

```{r}
gezondheidsmonitor.data <- gezondheidsmonitor.data %>%
  mutate(bmi = ifelse(bmi > 50, NA, bmi)) # Filter in main
```

# Clean Alcohol

[Placeholder for Text]

## alcohol_status and drinker_last_year

[Placeholder for Text]

```{r} 
gezondheidsmonitor.data <- gezondheidsmonitor.data %>%
  mutate(
    neverdrinker = case_when(
      drinker_last_year == 0 & exdrinker == 0 ~ 1,
      drinker_last_year == 1 & exdrinker == 0  ~ 0, 
      drinker_last_year == 0 & exdrinker == 1 ~ 0 
    )
  )
``` 

[Placeholder for Text]

```{r} 
gezondheidsmonitor.data <- gezondheidsmonitor.data %>% 
  mutate(
    alcohol_status = case_when(
      neverdrinker == 1 ~ 1, 
      drinker_last_year == 1 ~ 2, 
      exdrinker == 1 ~ 3
    )
  )

``` 

[Placeholder for Text]

```{r} 
gezondheidsmonitor.data <- gezondheidsmonitor.data %>% 
  mutate(
    drinker_last_year = case_when(
      neverdrinker == 1 ~ 0, 
      drinker_last_year == 1 ~ 1, 
      exdrinker == 1 ~ 0
    ), 
    drinker_last_year = as.integer(drinker_last_year) 
  )

``` 

## glasses_per_week

Non-drinkers have NA for the number of glasses in the data. We replace this with zero glasses. 

```{r} 
gezondheidsmonitor.data <- gezondheidsmonitor.data %>% 
  mutate(
    glasses_per_week = if_else(drinker_last_year == 1, glasses_per_week, 0)
  )
```

## drinker_over_a_year

```{r} 
gezondheidsmonitor.data <- gezondheidsmonitor.data %>% 
  mutate(
    drinker_over_a_week = case_when(
      glasses_per_week == 0 ~ 0, 
      glasses_per_week > 0 ~ 1
    ), 
    drinker_over_a_week = as.integer(drinker_over_a_week)
  )
```

## Summary

```{r} 
gezondheidsmonitor.data %>% 
  count(year, drinker_last_year, drinker_over_a_week, exdrinker, neverdrinker, alcohol_status)
``` 

# Explore Data

We will examine each variable in the dataset, now that the data have been cleaned. But let's first look at an overview of the dataset.

```{r}
vtable(gezondheidsmonitor.data, out = "return", missing = TRUE)
```

## Participation

[Placeholder for Text]

Let's see how many participants we have per year.

```{r}
gezondheidsmonitor.participants.by.year <- gezondheidsmonitor.data %>%
  group_by(year) %>%
  count(year)
gezondheidsmonitor.participants.by.year
```

We can visualise this in a plot.

```{r}
ggplot() +
  geom_col(
    mapping = aes(
      x = year,
      y = n,
      fill = factor(year)
    ),
    data = gezondheidsmonitor.participants.by.year,
    show.legend = FALSE
  ) +
  labs(
    x = "Year",
    y = "Number of participants"
  )
```

## Sex

Let's check, per year, what the number and percentages of males and females is.

```{r}
gezondheidsmonitor.prevalence.sex <- gezondheidsmonitor.data %>%
  count(year, sex) %>%
  group_by(year) %>%
  mutate(
    prevalence = n / sum(n)
  ) %>%
  ungroup()
gezondheidsmonitor.prevalence.sex
```

We can visualise this in a plot.

```{r}
ggplot() +
  geom_col(
    mapping = aes(
      x = factor(year),
      y = prevalence,
      fill = sex
    ),
    data = gezondheidsmonitor.prevalence.sex
  ) +
  labs(
    x = "Year",
    y = "Prevalence"
  ) +
  scale_fill_discrete(
    name = "Sex",
    label = c("0" = "Male", "1" = "Female")
  )
```

## Education

Let's check, per year, what the number and percentages of each of the three levels of education is.

```{r}
gezondheidsmonitor.prevalence.education <- gezondheidsmonitor.data %>%
  count(year, education) %>%
  group_by(year) %>%
  mutate(
    prevalence = n / sum(n)
  ) %>%
  ungroup()
gezondheidsmonitor.prevalence.education
```

We can visualise this in a plot.

```{r}
ggplot() +
  geom_col(
    mapping = aes(
      x = factor(year),
      y = prevalence,
      fill = education
    ),
    data = gezondheidsmonitor.prevalence.education
  ) +
  labs(
    x = "Year",
    y = "Prevalence"
  ) +
  scale_fill_discrete(
    name = "Level of education",
    labels = c("1" = "Low", "2" = "Medium", "3" = "High")
  )
```

## Age

Let's check, per year, what the minimum, mean and maximum age is.

```{r}
gezondheidsmonitor.data %>%
  group_by(year) %>%
  summarise(
    min.age = min(age),
    mean.age = mean(age),
    max.age = max(age),
    .groups = "drop"
  )
```

We can visualise the age distribution in a boxplot.

```{r}
ggplot() +
  geom_boxplot(
    mapping = aes(
      x = year,
      y = age,
      fill = factor(year)
    ),
    data = gezondheidsmonitor.data,
    show.legend = FALSE
  ) +
  labs(
    x = "Year",
    y = "Age"
  )
```

We can also visualise the probability density of the age distribution by year.

```{r}
ggplot() + 
  geom_density(
    mapping = aes(
      x = age,
      col = factor(year)
    ),
    data = gezondheidsmonitor.data
  ) +
  labs(
    x = "Age",
    y = "Probability density",
    col = "Year"
  )
```

## BMI

Let's check, per year, what the minimum, mean and maximum BMI value is.

```{r}
gezondheidsmonitor.data %>%
  filter(!is.na(bmi)) %>% # Remove in main
  group_by(year) %>%
  summarise(
    min.bmi = min(bmi),
    mean.bmi = mean(bmi),
    max.bmi = max(bmi),
    .groups = "drop"
  )
```

We can visualise the BMI distribution in a boxplot.

```{r}
ggplot() +
  geom_boxplot(
    mapping = aes(
      x = year,
      y = bmi,
      fill = factor(year)
    ),
    data = gezondheidsmonitor.data %>% filter(!is.na(bmi)), # Remove in main
    show.legend = FALSE
  ) +
  labs(
    x = "Year",
    y = "BMI"
  )
```

We can also visualise the probability density of the BMI distribution by year and sex.

```{r}
ggplot() +
  geom_density(
    mapping = aes(
      x = bmi,
      colour = sex,
      linetype = factor(year)
    ),
    data = gezondheidsmonitor.data %>%
      filter(
        bmi > 15,
        bmi < 45
      )
  ) +
  xlim(15, 45) +
  labs(
    x = "BMI",
    y = "Probability density",
    linetype = "Year"
  ) +
  scale_color_discrete(
    name = "Sex",
    labels = c("0" = "Male", "1" = "Female")
  )
```

# Write Output

Finally, we write the cleaned data of the Gezondheidsmonitor to a CSV file.

```{r}
write_csv(
  x = gezondheidsmonitor.data,
  file = "../../Input Data/Risk Factors/Gezondheidsmonitor.csv"
)
```

# References

<div id="refs"></div>
